---
title: "Arsenic Buccal DVP"
author: "Waverly Wei and Philippe Boileau"
output:
  html_document:
    toc: true
    toc_float: true
---

## Required Packages

```{r utilitylib, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(minfi)
library(limma)
library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
library(DMRcate)
library(doParallel)
library(GenomicFeatures)
library(org.Hs.eg.db)
library(missMethyl)
library(omicsPrint)
library(kableExtra)

# call helpers
source("../../helper-scripts/DVP-utility-functions.R")
source("../../helper-scripts/plotting-functions.R")
source("../../helper-scripts/probe-annotation.R")
source("../../helper-scripts/shared-helpers.R")

# knitr settings
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## Load Data

```{r data, cache=TRUE}
load(
  "../../../arsenic-epigenetics-meta/Chile/pbmcs/data/pbmcs_funnorm_data.RData"
)

# load cell counts
load(
  file = "../../../arsenic-epigenetics-meta/Chile/pbmcs/data/cell_counts_pbmcs.RData"
)

# add cell_counts to pheno
pheno <- cbind(pheno, cell_counts$counts)

# transform to working dataframe object
pheno <- as.data.frame(pheno)
```

## Exposure Counts

```{r exposure-dist-plot, echo=FALSE}
# Look at exposure distribution
exp_dist_p <- ggplot(pheno, aes(exposed)) +
  geom_histogram(alpha = 0.7, stat = "count", binwidth = 50) +
  xlab("Exposed?") +
  ylab("Count") +
  ggtitle("Distribution of Arsenic Exposure") +
  theme_minimal()

ggsave(filename = "DVP-plots/arsenic-exp-dist.png", device = "png", dpi = 300)
```
```{r, out.width = '100%'}
exp_dist_p
```

## Fit Models

```{r fit-models, cache=TRUE}
# Unadjusted model matrix
modUnadj <- model.matrix(~pheno$exposed)
                 
# Adjusted for cell type
modCell <- model.matrix(~pheno$exposed + pheno$CD8T + pheno$CD4T +
                         pheno$NK + pheno$Bcell + pheno$Mono + pheno$Gran)

# Adjusted for age, smoking, and sex
modAgeSmSex <- model.matrix(~pheno$exposed + pheno$age + pheno$smoking +
                              pheno$sex)
                   
# Adjusted for age, smoking, and sex
modCellAgeSmSex <- model.matrix(~pheno$exposed + pheno$age + pheno$smoking +
                                pheno$sex + pheno$CD8T + pheno$CD4T + pheno$NK +
                                pheno$Bcell + pheno$Mono + pheno$Gran)

# Run DVP functions on models
DVP_u <- run_DVP(mvals = mvals_fun, design = modUnadj)
DVP_cell <- run_DVP(mvals = mvals_fun, design = modCell)
DVP_ageSmSex <- run_DVP(mvals = mvals_fun, design = modAgeSmSex)
DVP_cellAgeSmSex <- run_DVP(mvals = mvals_fun, design = modCellAgeSmSex)
```

## DVP Results 

The results are compilued using the Bonferroni correction, but we would be
better off using Holm's procedure. It is uniformely more powerful than the
Bonferroni correction.

```{r dvpRes, cache=TRUE}
# Unadjusted model ###########

# filter by p values
table(DVP_u$P.Value < 0.05)
# filter by BH-adjusted-pvalues
table(DVP_u$Adj.P.Value < 0.05)
# filter by Bonferroni-adjusted pvalues
table(DVP_u$Bonferroni < 0.05)

# save results
write.table(
  DVP_u,
  file = "DVP-results/unadjusted.txt",
  sep = "\t",
  row.names = FALSE,
  col.names = TRUE
)

# Adjusted for cell ###########

# filter by p values
table(DVP_cell$P.Value < 0.05)
# filter by BH-adjusted-pvalues
table(DVP_cell$Adj.P.Value < 0.05)
# filter by Bonferroni-adjusted pvalues
table(DVP_cell$Bonferroni < 0.05)

# save results
write.table(
  DVP_cell,
  file = "DVP-results/cell.txt",
  sep = "\t",
  row.names = FALSE,
  col.names = TRUE
)


# Adjusted for age, sex, and smoking status

# filter by p values
table(DVP_ageSmSex$P.Value < 0.05)
# filter by BH-adjusted-pvalues
table(DVP_ageSmSex$Adj.P.Value < 0.05)
# filter by Bonferroni-adjusted pvalues
table(DVP_ageSmSex$Bonferroni < 0.05)

# save results
write.table(
  DVP_ageSmSex,
  file = "DVP-results/age-sex-smoking.txt",
  sep = "\t",
  row.names = FALSE,
  col.names = TRUE
)

# Adjusted for cell, age, sex, and smoking status

# filter by p values
table(DVP_cellAgeSmSex$P.Value < 0.05)
# filter by BH-adjusted-pvalues
table(DVP_cellAgeSmSex$Adj.P.Value < 0.05)
# filter by Bonferroni-adjusted pvalues
table(DVP_cellAgeSmSex$Bonferroni < 0.05)

# save results
write.table(
  DVP_cellAgeSmSex,
  file = "DVP-results/cell-age-sex-smoking.txt",
  sep = "\t",
  row.names = FALSE,
  col.names = TRUE
)
```


## Annotate Probes

We now annotate the probes that are significant (after the Bonferroni
correction) in the full model.

### Full Model

```{r full-model-qq, cache=TRUE}
# create the qqplot for the full model
gg_qqplot(DVP_cellAgeSmSex$P.Value)
```

```{r annotDVP-full, cache=TRUE}
# annotate full model results
DVP_sig <- DVP_cellAgeSmSex %>%
  filter(Bonferroni < 0.05)

DVP_annot <- annotate_probes(probeID_column = 1, DVP_sig)

data(hm450.manifest.pop.GoNL)
dat <- data.frame(hm450.manifest.pop.GoNL)
rownames(dat) <- names(hm450.manifest.pop.GoNL)
maskname <- rownames(dat)[
  which(dat[, paste("MASK.general.", "EAS", sep = "")] == TRUE)]

DVP_CpG <- DVP_annot$Probe_ID
length(unique(DVP_CpG[(DVP_CpG %in% maskname)]))

DVP_annot <- DVP_annot[!(DVP_annot$Probe_ID %in% maskname), ]
bonf <- dplyr::filter(DVP_annot, Bonferroni < 0.05)
DVP_bonf <- bonf$Probe_ID
length(unique(DVP_bonf[(DVP_bonf %in% maskname)]))

DVP_annot <- DVP_annot %>% arrange(Bonferroni)
```

```{r annotDVPres}
DVP_annot %>%
  kbl() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")
```

### Almost Full Model (No Cells)

```{r no-cell-qq, cache=TRUE}
# qqplot of model with no cells
gg_qqplot(DVP_ageSmSex$P.Value)
```

```{r annotDVP-no-cells, cache=TRUE}
# annotate model withouth cells
DVP_sig <- DVP_ageSmSex %>%
  filter(Bonferroni < 0.05)

DVP_annot <- annotate_probes(probeID_column = 1, DVP_sig)
DVP_CpG <- DVP_annot$Probe_ID
length(unique(DVP_CpG[(DVP_CpG %in% maskname)]))

DVP_annot <- DVP_annot[!(DVP_annot$Probe_ID %in% maskname), ]
bonf <- dplyr::filter(DVP_annot, Bonferroni < 0.05)
DVP_bonf <- bonf$Probe_ID
length(unique(DVP_bonf[(DVP_bonf %in% maskname)]))

DVP_annot <- DVP_annot %>% arrange(Bonferroni)
```

```{r annotDVPres-no-cells}
DVP_annot %>%
  kbl() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")
```
