---
title: "DMR Meta-Analysis"
author: "Philippe Boileau"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(kableExtra)

# call helpers
source("../helper-scripts/probe-annotation.R")

# knitr settings
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Create BED Files

Before running the meta-analysis results through `combined-pvalues`, we must
format them into BED files. We therefore begin by annotating the probes
described in the DMP and DVP meta-analysis result files.

```{r get-annotated-data, eval=FALSE}
# load in the meta analysis results

# DMPs
DMP_all_cell_common <- read_csv(
  "data/DMPs/DMP_all_cell_common_120420.csv"
) %>% 
  annotate_probes_DMR(annotation)
DMP_all_no_cell_common <- read_csv(
  "data/DMPs/DMP_all_no_cell_common_120420.csv"
) %>%
  annotate_probes_DMR(annotation)
DMP_blood_cell_common <- read_csv(
  "data/DMPs/DMP_blood_cell_common_120420.csv"
) %>%
  annotate_probes_DMR(annotation)
DMP_blood_no_cell_common <- read_csv(
  "data/DMPs/DMP_blood_no_cell_common_120420.csv"
) %>%
  annotate_probes_DMR(annotation)

# DVPs
DVP_all_cell_common <- read_csv(
  "data/DVPs/DVP_all_cell_common_120420.csv"
) %>%
  annotate_probes_DMR(annotation)
DVP_all_no_cell_common <- read_csv(
  "data/DVPs/DVP_all_no_cell_common_120420.csv"
) %>%
  annotate_probes_DMR(annotation)
DVP_blood_cell_common <- read_csv(
  "data/DVPs/DVP_blood_cell_common_120420.csv"
) %>%
  annotate_probes_DMR(annotation)
DVP_blood_no_cell_common <- read_csv(
  "data/DVPs/DVP_blood_no_cell_common_120420.csv"
) %>%
  annotate_probes_DMR(annotation)
```

Now that the probes have been annotated, we can write the BED files.

```{r create-BED-files, eval=FALSE}
# DMPs
DMP_all_cell_common %>% write_tsv(
  path = "data/DMPs/DMP_all_cell_common.bed"
)
DMP_all_no_cell_common %>% write_tsv(
  path = "data/DMPs/DMP_all_no_cell_common.bed"
)
DMP_blood_cell_common %>% write_tsv(
  path = "data/DMPs/DMP_blood_cell_common.bed"
)
DMP_blood_no_cell_common %>% write_tsv(
  path = "data/DMPs/DMP_blood_no_cell_common.bed"
)

# DVPs
DVP_all_cell_common %>% write_tsv(
  path = "data/DVPs/DVP_all_cell_common.bed"
)
DVP_all_no_cell_common %>% write_tsv(
  path = "data/DVPs/DVP_all_no_cell_common.bed"
)
DVP_blood_cell_common %>% write_tsv(
  path = "data/DVPs/DVP_blood_cell_common.bed"
)
DVP_blood_no_cell_common %>% write_tsv(
  path = "data/DVPs/DVP_blood_no_cell_common.bed"
)
```


# `combined-pvalues`

Now that the BED files for each meta-analysis have been created, we run them
through the `combined-pvalues` Python library's `comb-p` command line function.
A brief introduction to this method can be found 
[here](https://github.com/brentp/combined-pvalues). Based on the findings of
[Peters et al.](https://epigeneticsandchromatin.biomedcentral.com/articles/10.1186/1756-8935-8-6), we set the `dist=1000`. An example of the command line code for running
`comb-p` can be found in the code chunk below.

```{bash, eval=FALSE}
comb-p pipeline \
  -c 6 \               # p-values stores in 6th column
  -p results/DMPs/...  # prefix of where results will be stores (DMPs or DVPs)
  --seed 1e-3 \        # require a p-value of 1e-3 to start a region
  --dist 1000 \        # extend region if find another p-value within this dist
  --region-filter-p 0.1 \         # post-filter reported regions
  --region-filter-n 3   \         # post-filter size of region
  --anno hg19                     # annotation information
  data/DMP_or_DVP_file_here.bed   # name of BED file
```


# Results

The DMR results are presented in the tables below.

## DMR Results for DMPs

### All Datasets, Controlling for Cells

```{r}
res_tbl <- read_tsv(
  "results/DMPs/all_cell_common/all_cell_common.anno.hg19.bed"
) %>%
  arrange(min_p)
res_tbl %>%
  kbl() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")
```
```{r include=FALSE, eval=FALSE}
res_tbl %>%
  write_csv(
    "results/DMPs/all_cell_common/all_cell_common.results.csv"
  )
```



### All Datasets, Not Controlling for Cells

```{r}
res_tbl <- read_tsv(
  "results/DMPs/all_no_cell_common/all_no_cell_common.anno.hg19.bed"
) %>%
  arrange(min_p)
res_tbl%>%
  kbl() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")
```
```{r include=FALSE, eval=FALSE}
res_tbl %>%
  write_csv(
    "results/DMPs/all_no_cell_common/all_no_cell_common.results.csv"
  )
```

### No Buccal Dataset, Controlling for Cells

```{r}
res_tbl <- read_tsv(
  "results/DMPs/blood_cell_common/blood_cell_common.anno.hg19.bed"
) %>%
  arrange(min_p)
res_tbl%>%
  kbl() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")
```
```{r include=FALSE, eval=FALSE}
res_tbl %>%
  write_csv(
    "results/DMPs/blood_cell_common/blood_cell_common.results.csv"
  )
```

### No Buccal Dataset, Not Controlling for Cells

```{r}
res_tbl <- read_tsv(
  "results/DMPs/blood_no_cell_common/blood_no_cell_common.anno.hg19.bed"
) %>%
  arrange(min_p)
res_tbl %>%
  kbl() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")
```
```{r include=FALSE, eval=FALSE}
res_tbl %>%
  write_csv(
    "results/DMPs/blood_no_cell_common/blood_no_cell_common.results.csv"
  )
```

## DMR Results for DVPs

### All Datasets, Controlling for Cells

```{r}
res_tbl <- read_tsv(
  "results/DVPs/all_cell_common/all_cell_common.anno.hg19.bed"
) %>%
  arrange(min_p)
res_tbl %>%
  kbl() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")
```
```{r include=FALSE, eval=FALSE}
res_tbl %>%
  write_csv(
    "results/DVPs/all_cell_common/all_cell_common.results.csv"
  )
```

### All Datasets, Not Controlling for Cells

```{r}
res_tbl <- read_tsv(
  "results/DVPs/all_no_cell_common/all_no_cell_common.anno.hg19.bed"
) %>%
  arrange(min_p)
res_tbl %>%
  kbl() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")
```
```{r include=FALSE, eval=FALSE}
res_tbl %>%
  write_csv(
    "results/DVPs/all_no_cell_common/all_no_cell_common.results.csv"
  )
```

### No Buccal Dataset, Controlling for Cells

```{r}
res_tbl <- read_tsv(
  "results/DVPs/blood_cell_common/blood_cell_common.anno.hg19.bed"
) %>%
  arrange(min_p)
res_tbl %>%
  kbl() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")
```
```{r include=FALSE, eval=FALSE}
res_tbl %>%
  write_csv(
    "results/DVPs/blood_cell_common/blood_cell_common.results.csv"
  )
```

### No Buccal Dataset, Not Controlling for Cells

```{r}
res_tbl <- read_tsv(
  "results/DVPs/blood_no_cell_common/blood_no_cell_common.anno.hg19.bed"
) %>%
  arrange(min_p)
res_tbl %>%
  kbl() %>%
  kable_styling() %>%
  scroll_box(width = "100%", height = "300px")
```
```{r include=FALSE, eval=FALSE}
res_tbl %>%
  write_csv(
    "results/DVPs/blood_no_cell_common/blood_no_cell_common.results.csv"
  )
```
