---
title: "Buccal Cell Preprocessing Pipeline -- Chile Arsenic Study"
author: "Waverly Wei and Philippe Boileau"
output:
  html_document:
    toc: true
    toc_float: true
---

## Required Packages

```{r library, message=FALSE, warning=FALSE}
library(tidyverse)
library(RColorBrewer)
library(readxl)
library(minfi)
library(ggplot2)
library(ENmix)
library(stringr)
library(geneplotter)
library(ChAMP)
library(wateRmelon)
library(sva)
library(pdftools)
library(filesstrings)
library(Harman)
library(DMRcate)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

## Prepare and Load Data

### Metadata Preparation

First, we load the buccal cell sample metadata. Note that the orignal
sample metadata had a number of errors: there was a duplication of sample IDs.

```{r data, eval=FALSE, echo=TRUE}
# Get the sample info. Note that the top-level file for this R project is a
# clone of the "Superfund-Methylatio" GitHub repository. The data is stored in a
# copy of the "Arsenic Epigeenetics META" repository, locally named
# "arsenic-epigenetics-meta".
sample_metadata <- read_xlsx(
  path = paste0(
    "../../../arsenic-epigenetics-meta/Chile/20170212_AC2013_DNAmethArray_Buccal",
    "_PBMCs_MetaData.xlsx"),
  sheet = "AC2013_DNAmeth_PL2_Buccal"
) %>%
  mutate(
    position = `Position (on FINAL PLATE)`,
    sample_id = `Sample ID`,
    buccal_sample_id = `BUCCAL Sample ID`,
    study_id = `Study_ID...5`,
    exposed = as.factor(if_else(`Exposed: Yes:1; No:2` == 1, "yes", "no")),
    age = P3Q8_Age,
    sex = as.factor(if_else(`P3Q9_Sex_(M=1,F=2)` == 1, "M", "F")),
    smoking = as.factor(if_else(P38Q96_Smoking == 1, "yes", "no")),
    secondhand_smoking = as.factor(
      if_else(P40Q116_Smoking_Secondhand_for_non_smokers == "1", "yes", "no")
    ),
    diabetes = as.factor(if_else(P11Q35_Diabetes == 1, "yes", "no")),
    skin_cancer = as.factor(if_else(P10Q33_Cancer_Skin == 1, "yes", "no")),
    hypertension = as.factor(if_else(P12Q37_Hypertension == 1, "yes", "no")),
    cardio = as.factor(if_else(P13Q38_Cardio == 1, "yes", "no")),
    breastfed = as.factor(if_else(P26Q63_Breastfed == "1", "yes", "no")),
    gastric = as.factor(if_else(P16Q43_Gastric == 1, "yes", "no")),
    tuberculosis = as.factor(if_else(P15Q42_Tuberculosis == 1, "yes", "no")),
    intestinal = as.factor(if_else(P17Q45_Intestinal == 1, "yes", "no")),
    childhood_diseases = as.factor(
      if_else(P20Q50_Childhood_diseases == 1, "yes", "no")
    ),
    anemia = as.factor(if_else(P18Q47_Anemia == 1, "yes", "no")),
    immune = as.factor(if_else(P19Q48_Immune == 1, "yes", "no")),
    antibiotics = as.factor(
      if_else(`P22Q54_Antibiotics_(Yes=1,No=2)` == 1, "yes", "no")
    ),
    immunosuppresor = as.factor(
      if_else(`P24Q58_Immuosupresores_(Yes=1,No=2)` == 1, "yes", "no")
    ),
    laxatives = as.factor(
      if_else(`P25Q62_laxatives_(Yes=1,No=2)` == 1, "yes", "no")
    ),
    med_high_triglycerides = as.factor(
      if_else(Med_High_triglycerides == 1, "yes", "no")
    ),
    med_insulin_resistance = as.factor(
      if_else(Med_Insulin_resistance == 1, "yes", "no")
    ),
    med_hypertension = as.factor(if_else(Med_Hypertension == 1, "yes", "no")),
    med_diabetes = as.factor(if_else(Med_Diabetes == 1, "yes", "no")),
    med_high_cholesterol = as.factor(
      if_else(Med_High_Cholesterol == 1, "yes", "no")
    ),
    med_anticoagulant = as.factor(if_else(Med_anticoagulant == 1, "yes", "no")),
    weight_20 = P4Q12a_Weight20,
    weight_40 = P4Q12b_Weight40
  ) %>%
  dplyr::select(
    position, sample_id, buccal_sample_id, study_id, exposed, age, sex, smoking,
    secondhand_smoking, diabetes, skin_cancer, hypertension, cardio, breastfed,
    gastric, tuberculosis, intestinal, childhood_diseases, anemia, immune,
    antibiotics, immunosuppresor, laxatives, med_high_triglycerides,
    med_insulin_resistance, med_hypertension, med_diabetes,
    med_high_cholesterol, med_anticoagulant, weight_20, weight_40
  ) %>%
  filter(
    sample_id != "QC", buccal_sample_id != "BLANK"
  )

# assign duplicates
sample_metadata <- sample_metadata %>%
  mutate(
    duplicates = if_else(sample_id %in% c("B1", "B36"), 1,
                   if_else(sample_id %in% c("B16", "B34"), 2,
                     if_else(sample_id %in% c("B2", "B5"), 3,
                       if_else(sample_id %in% c("B7", "B41"), 4, 0))))
  )

# now load the sentrix id and position data
sample_info <- read_csv(
  "../../../arsenic-epigenetics-meta/Chile/Sample_Info-Berkeley.csv"
) %>%
  dplyr::select(
    Sample.ID, Sample_Well, Sentrix_ID, Sentrix_Position
  ) %>%
  mutate(
    sentrix_full = paste(Sentrix_ID, Sentrix_Position, sep = "_"),
    sample_id = Sample.ID,
    position = Sample_Well
  ) %>%
  dplyr::select(
    sample_id, position, sentrix_full
  ) %>%
  filter(
    sample_id %in% paste0("B", 1:47)
  )
 
# create a vector of idat filenames
sample_info$Basename <- paste0(
  paste0("/Users/philippe/Documents/berkeley/",
         "superfund/arsenic-epigenetics-meta/Chile/idat/"),
  sample_info$sentrix_full
)

# include the base idat filenames to the sample metadata
sample_info <- sample_info %>%
  dplyr::select(position, Basename)
sample_metadata <- sample_metadata %>%
  left_join(sample_info, by = "position")
```

With the meta data completed, we now remove the duplicates from the data,
retaining only the first sample in order of sample ID from each individual. The 
duplicated samples are listed below:

- Participant AC130004: B1 and B36
- Participant AC130037: B16 and B34
- Participant AC130044: B2 and B5
- Participant AC130079: B7 and B41

```{r rm-duplicates, echo=TRUE, eval=FALSE}
keep <- !(sample_metadata$sample_id %in% c("B36", "B34", "B5", "B41"))
sample_metadata <- sample_metadata[keep, ]
rm(sample_info, keep)
```


### Create `RGChannelSetExtended` Object

Next, we create or load the `RGChannelSetExtended` object, depending on whether
or not it's been generated locally in the past.

```{r RGset, eval=FALSE}
# Change to eval = TRUE if RGset dataset does not exist in the environment. Plug
# in sample metadata generated from the previous code chunk.
RGset <- read.metharray.exp(targets = sample_metadata, extended = TRUE)
save(
  RGset,
  file = "../../../arsenic-epigenetics-meta/Chile/buccal/data/buccal_RGset.RData"
)
```

```{r pheno, echo=TRUE}
# if the RGset already exists, run this code chunk instead of the one above
load("../../../arsenic-epigenetics-meta/Chile/buccal/data/buccal_RGset.RData")
pheno <- pData(RGset)
sample_name <- rownames(pheno)
pheno <- data.frame(pheno, sample_name)

# adding Sentrix ID for subsequent batch correction in downstream analysis
sentrix_id <- as.factor(
  stringr::str_extract_all(pheno$sample_name, "^([0-9]{12})",
                           simplify = TRUE)
)
pheno <- data.frame(pheno, sentrix_id)

# remove features without any variation
pheno <- pheno[, apply(pheno, 2, function(x) length(unique(x)) > 1)]
```

## Internal Quality Control

First, we check that the signal to noise ratio of each sample. Sample B14's
detection p-values are significantly larger than the rest, and so we remove it
from consideration.

```{r detection_p-values, echo=TRUE}
# check detection p-values
det_p <- detectionP(RGset)
pal <- brewer.pal(8, "Dark2")
barplot(
  colMeans(det_p), col = pal[factor(pheno$sample_id)], las = 2,
  cex.names = 0.8, ylab = "Mean detection p-values",
  names.arg = pheno$sample_id
)

# remove B14
RGset <- RGset[, -26]
pheno <- pData(RGset)

det_p <- detectionP(RGset)
pal <- brewer.pal(8, "Dark2")
barplot(
  colMeans(det_p), col = pal[factor(pheno$sample_id)], las = 2,
  cex.names = 0.8, ylab = "Mean detection p-values",
  names.arg = pheno$sample_id
)
```

We then produce a PDF version QC report for Illumina Infinium Human Methylation
450k arrays, which is useful for identifying failed samples.

```{r internal-control, eval=FALSE}
qcReport(
  RGset,
  sampGroups = pheno$exposed,
  pdf = "preprocessing-reports/buccal_ControlProbes.pdf"
)
```

## Outlier information

### Pre-Filtering

A number of functions are run sequentially on the `RGset` object. First,
outliers are identified based on their median unmethylated/methylated intensity.

```{r outlier-information, include=FALSE, echo=TRUE}
Mset <- preprocessRaw(RGset)
QC_minfi <- minfiQC(Mset, fixOutliers = TRUE)
```

```{r save-outlier-plot, eval=FALSE}
pdf(file = "preprocessing-reports/buccal_PotentialOutliers.pdf")
plotQC(QC_minfi$qc)
dev.off()
```

```{r outlierplot, out.width='100%'}
plotQC(QC_minfi$qc)
```

Next, `champ.QC` is used to perform a brief exploratory data analysis. The
resulting plots are listed below.

```{r prefiltering-eda, eval=FALSE, echo=TRUE}
# Create pdf of quality control plots: mdsplot, densityPlot, dendrogram, et.c
champ.QC(
  beta = getBeta(Mset), pheno = pheno$exposed, mdsPlot = TRUE,
  densityPlot = TRUE, dendrogram = TRUE, PDFplot = TRUE, Rplot = FALSE,
  Feature.sel = "None", resultsDir = "preprocessing-reports/QC_preFiltering"
)
```
```{r, eval=FALSE}
#  convert pdfs to pngs for printing
pdf_convert(
  pdf = "preprocessing-reports/QC_preFiltering/raw_densityPlot.pdf",
  format = "png",
  dpi = 300,
  filenames = "preprocessing-reports/QC_preFiltering/raw_densityPlot.png"
)
pdf_convert(
  pdf = "preprocessing-reports/QC_preFiltering/raw_mdsPlot.pdf",
  format = "png",
  dpi = 300,
  filenames = "preprocessing-reports/QC_preFiltering/raw_mdsPlot.png"
)
pdf_convert(
  pdf = "preprocessing-reports/QC_preFiltering/raw_SampleCluster.pdf",
  format = "png",
  dpi = 300,
  filenames = "preprocessing-reports/QC_preFiltering/raw_SampleClustert.png"
)
```

Pre-Filtering Beta Values

```{r prebeta, out.width='100%'}
knitr::include_graphics(
  "preprocessing-reports/QC_preFiltering/raw_densityPlot.png"
)
```

Pre-Filtering MDS Plot

```{r premds, out.width='100%'}
knitr::include_graphics(
  "preprocessing-reports/QC_preFiltering/raw_mdsPlot.png"
)
```

Pre-Filtering Dendogram

```{r predendo, out.width='100%'}
knitr::include_graphics(
  "preprocessing-reports/QC_preFiltering/raw_SampleClustert.png"
)
```

Pre-Filtering SVD

```{r preSVD, eval=FALSE, echo=TRUE}
# remove uninteresting variables from SVD
rm_svd <- c("position", "sample_id", "buccal_sample_id", "study_id",
            "duplicates", "Basename", "filenames", "sample_name", "sentrix_id",
            "xMed", "yMed", "predictedSex")

# run SVD
champ.SVD(
  beta = getBeta(Mset)[complete.cases(getBeta(Mset)), ],
  pd = pheno[, -which(colnames(pheno) %in% rm_svd)],
  RGEffect = TRUE, Rplot = FALSE,
  resultsDir = "preprocessing-reports/QC_preFiltering/"
)
```
```{r, eval=FALSE}
# save the plots
pdf_convert(
  pdf = "preprocessing-reports/QC_preFiltering/SVDsummary.pdf",
  format = "png",
  dpi = 300,
  filenames = c("preprocessing-reports/QC_preFiltering/raw_svd_loadings.png",
                "preprocessing-reports/QC_preFiltering/raw_screeplot.png")
)
```

```{r presvd, out.width='100%'}
knitr::include_graphics(
  "preprocessing-reports/QC_preFiltering/raw_svd_loadings.png"
)
knitr::include_graphics(
  "preprocessing-reports/QC_preFiltering/raw_screeplot.png"
)
```


## Post Filtering

Finally, quantitiative measures of data quality are computed, including
number of beads, averaged bisulfite conversion intensity, and detection
p-values of probes. Low quality samples and probes are identified, as well as
the outliers based on total intensity and beta value distributions. The results
are listed below.

```{r qc-outlier-detection, echo=TRUE}
# compute QC metrics
qc <- ENmix::QCinfo(RGset, distplot = FALSE)
```
```{r move-files, eval=FALSE}
# move generated plots to appropriate directory
file.move(
  "qc_CpG.jpg",
  "preprocessing-reports/QC_preFiltering/",
  overwrite = TRUE
)
file.move(
  "qc_sample.jpg",
  "preprocessing-reports/QC_preFiltering/",
  overwrite = TRUE
)
```

```{r qc-filter-samples-and-probes, echo=TRUE}
# beta: percentage of metylation
# M: log2 ratio of intensities of methylated vs. unmethylated
betas <- rmSNPandCH(
  getBeta(Mset), dist = 2, mafcut = 0.05, and = TRUE, rmcrosshyb = TRUE,
  rmXY = FALSE
)
mvals <- beta2m(betas)

# removing filtered probes from qc
qc$nbead <- qc$nbead[rownames(qc$nbead) %in% rownames(betas), ]
qc$detP <- qc$detP[rownames(qc$detP) %in% rownames(betas), ]

filtered <- champ.filter(
  beta = betas, M = mvals,
  pd = pheno, beadcount = qc$nbead, detP = qc$detP,
  arraytype = "EPIC", filterXY = TRUE, SampleCutoff = 0.05
)
```

We next perform the same exploratory data analysis as before, but with the
low-quality probes removed.

```{r filter-samples-and-probes, echo=TRUE, eval=FALSE}
# filter sample and probes
pheno <- filtered$pd
betas <- filtered$beta
mvals <- filtered$M

filtered_CpG <- rownames(betas)
unfiltered_CpG <- rownames(getBeta(Mset))
outCpG <- unfiltered_CpG[!(unfiltered_CpG %in% filtered_CpG)]
  
# recompute the density plots
champ.QC(
  beta = betas, pheno = pheno$exposed, mdsPlot = TRUE,
  densityPlot = TRUE, dendrogram = TRUE, PDFplot = TRUE, Rplot = FALSE,
  Feature.sel = "None",
  resultsDir = "preprocessing-reports/QC_postFiltering"
)

# save the csv
write.csv(
  pheno,
  file = "../../../arsenic-epigenetics-meta/Chile/buccal/data/buccal_pheno.csv"
)
```
```{r, eval=FALSE}
#  convert pdfs to pngs for printing
pdf_convert(
  pdf = "preprocessing-reports/QC_postFiltering/raw_densityPlot.pdf",
  format = "png",
  dpi = 300,
  filenames = "preprocessing-reports/QC_postFiltering/raw_densityPlot.png"
)
pdf_convert(
  pdf = "preprocessing-reports/QC_postFiltering/raw_mdsPlot.pdf",
  format = "png",
  dpi = 300,
  filenames = "preprocessing-reports/QC_postFiltering/raw_mdsPlot.png"
)
pdf_convert(
  pdf = "preprocessing-reports/QC_postFiltering/raw_SampleCluster.pdf",
  format = "png",
  dpi = 300,
  filenames = "preprocessing-reports/QC_postFiltering/raw_SampleClustert.png"
)
```

Post-Filtering Beta Values

```{r post-beta, out.width='100%'}
knitr::include_graphics(
  "preprocessing-reports/QC_postFiltering/raw_densityPlot.png"
)
```

Post-Filtering MDS Plot

```{r post-mds, out.width='100%'}
knitr::include_graphics(
  "preprocessing-reports/QC_postFiltering/raw_mdsPlot.png"
)
```

Post-Filtering Dendogram

```{r post-dendo, out.width='100%'}
knitr::include_graphics(
  "preprocessing-reports/QC_postFiltering/raw_SampleClustert.png"
)
```

Post-Filtering SVD

```{r post-SVD, eval=FALSE, echo=TRUE}
# run SVD
champ.SVD(
  beta = betas,
  pd = pheno[, -which(colnames(pheno) %in% rm_svd)],
  RGEffect = TRUE, Rplot = FALSE,
  resultsDir = "preprocessing-reports/QC_postFiltering/"
)
```
```{r, eval=FALSE}
# save the plots
pdf_convert(
  pdf = "preprocessing-reports/QC_postFiltering/SVDsummary.pdf",
  format = "png",
  dpi = 300,
  filenames = c("preprocessing-reports/QC_postFiltering/raw_svd_loadings.png",
                "preprocessing-reports/QC_postFiltering/raw_screeplot.png")
)
```

```{r post-svd-plots, out.width='100%'}
knitr::include_graphics(
  "preprocessing-reports/QC_postFiltering/raw_svd_loadings.png"
)
knitr::include_graphics(
  "preprocessing-reports/QC_postFiltering/raw_screeplot.png"
)
```


## Normalization

We now normalize the data, and re-perform the exploratory data analysis to
validate that the quality control measures were successful.

### Funnorm Normalization

```{r fun-normalization, eval=FALSE, echo=TRUE}
funnorm <- preprocessFunnorm(RGset)
betas_fun <- minfi::getBeta(funnorm)
betas_fun <- betas_fun[rownames(betas_fun) %in% rownames(filtered$beta), ]
mvals_fun <- B2M(betas_fun)
pheno <- pData(funnorm)

# save the processed beta and mvals, as well as pheno data
save(
  betas_fun, mvals_fun, pheno,
  file = "../../../arsenic-epigenetics-meta/Chile/buccal/data/buccal_funnorm_data.RData"
)
```
```{r post-filter-norm-eda, eval=FALSE, echo=TRUE}
# perform EDA
champ.QC(
  beta = betas_fun, pheno = pheno$exposed, mdsPlot = TRUE, densityPlot = TRUE,
  dendrogram = TRUE, PDFplot = TRUE, Rplot = FALSE, Feature.sel = "None",
  resultsDir = "preprocessing-reports/QC_postNormalization/"
)

champ.SVD(
  beta = betas_fun, pd = pheno[, -which(colnames(pheno) %in% rm_svd)],
  RGEffect = TRUE, Rplot = FALSE,
  resultsDir = "preprocessing-reports/QC_postNormalization/"
)
```
```{r, eval=FALSE}
#  convert pdfs to pngs for printing
pdf_convert(
  pdf = "preprocessing-reports/QC_postNormalization/raw_densityPlot.pdf",
  format = "png",
  dpi = 300,
  filenames = "preprocessing-reports/QC_postNormalization/raw_densityPlot.png"
)
pdf_convert(
  pdf = "preprocessing-reports/QC_postNormalization/raw_mdsPlot.pdf",
  format = "png",
  dpi = 300,
  filenames = "preprocessing-reports/QC_postNormalization/raw_mdsPlot.png"
)
pdf_convert(
  pdf = "preprocessing-reports/QC_postNormalization/raw_SampleCluster.pdf",
  format = "png",
  dpi = 300,
  filenames = "preprocessing-reports/QC_postNormalization/raw_SampleClustert.png"
)
pdf_convert(
  pdf = "preprocessing-reports/QC_postNormalization/SVDsummary.pdf",
  format = "png",
  dpi = 300,
  filenames = c(
    "preprocessing-reports/QC_postNormalization/raw_svd_loadings.png",
    "preprocessing-reports/QC_postNormalization/raw_screeplot.png"
  )
)
```

Funnorm Normalization Betas Distribution

```{r funnorm-beta, out.width='100%'}
knitr::include_graphics(
  "preprocessing-reports/QC_postNormalization/raw_densityPlot.png"
)
```

Funnorm Normalization MDS Plot

```{r funnorm-mds, out.width='100%'}
knitr::include_graphics(
  "preprocessing-reports/QC_postNormalization/raw_mdsPlot.png"
)
```

Funnorm Normalization Dendogram

```{r funnormdendo, out.width='100%'}
knitr::include_graphics(
  "preprocessing-reports/QC_postNormalization/raw_SampleClustert.png"
)
```


SVD Results

```{r post-norm-svd, out.width='100%'}
knitr::include_graphics(
  "preprocessing-reports/QC_postNormalization/raw_svd_loadings.png"
)
knitr::include_graphics(
  "preprocessing-reports/QC_postNormalization/raw_screeplot.png"
)
```

## Estimate Cell Counts

We now estimate the cell-type decomposition of each sample, and plot their
distributions. The cell types are estimated via ReFACTor, a method based on
sparse principal componenent analysis and designed for cell type heterogeneity
in EWAS. It provides the estimates of cell type composisition. The bash script
for generating these counts is provided below. For installation instructions,
see the [ReFACTor GitHub repo](https://github.com/cozygene/refactor).

```{r prep-cell-counts, eval=FALSE, echo=TRUE}
# create betas tsv
betaRefact <- cbind(rownames(betas_fun), data.frame(betas_fun))
colnames(betaRefact) <- str_sub(colnames(betaRefact), start = 2L)
colnames(betaRefact)[1] <- "ID"
write.table(betaRefact, file = 'preprocessing-reports/betas_ReFACTor.txt',
            sep = "\t", col.names = TRUE, row.names = FALSE)

# create pheno tsv
pheno$ID <- rownames(pheno)
subset_pheno <- pheno[, c("ID", "sex", "smoking")]
subset_pheno <- subset_pheno %>%
  as_tibble %>%
  mutate(
    sex = if_else(sex == "M", 0, 1),
    smoking = if_else(smoking == "yes", 1, 0)
  )
write.table(subset_pheno, file = "preprocessing-reports/pheno_ReFACTor.txt",
            sep = "\t", col.names = FALSE, row.names = FALSE)
```
```{bash generate-cell-counts, eval=FALSE, echo=TRUE}
# make sure that you're in the right directory
cd  superfund/Superfund-Methylation/Arsenic-buccal/preprocessing-reports

# estimate the number of sPCs to use (note that we can't take into account the
# effect of controlling for smoking and sex).
python estimate_k.py --datafile betas_ReFACTor.txt

# run ReFACTor (might need to run w/ pythonw if using conda)
python refactor.py --datafile betas_ReFACTor.txt --k 6 \
  --covarfile pheno_ReFACTor.txt \
  --out preprocessing-reports/refactor_buccal
```
