---
title: "Arsenic and DNA Methylation Analysis -- Table 1, Chile"
author: "Philippe Boileau"
date: "8/31/2020"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(knitr)
library(readxl)

knitr::opts_chunk$set(echo = FALSE)
```

## Identifying Technical Replicates

*This first section has been kept for posterity. Review the second section for up-to-date descriptive statistics of the Chile study's participants.*

The Chilean participants' characteristics are summarized in the tables below.
The participant's metadata dataframes stored in the buccal and PBMC folders were
used independently to determine if there were any differences.

```{r load-data, warning=FALSE, message=FALSE}
buccal_pheno_tbl <- read_csv(
  "../../../arsenic-epigenetics-meta/Chile/buccal/data/buccal_pheno_og.csv"
)
pbmc_pheno_tbl <- read_csv(
  "../../../arsenic-epigenetics-meta/Chile/pbmcs/data/pbmcs_20200506_pheno.csv"
)
```

```{r create-table-buccal, warning=FALSE, message=FALSE}
buccal_table <- buccal_pheno_tbl %>%
  mutate(
    male = if_else(P3Q9_Sex == "a", 1, 0), # assume that a == man, b == women
    ever_smoker = if_else(P38Q96_Smoking == "a", 1, 0),
    prenatal_exp = if_else(Exposed..Yes.a..No.b == "a", 1, 0)
  )  %>%
  summarise(
    participant_count = n(),
    mean_age = mean(P3Q8_Age),
    sd_age = sd(P3Q8_Age),
    count_male = sum(male),
    prop_male = mean(male),
    count_ever_smoker = sum(ever_smoker),
    prop_ever_smoker = mean(ever_smoker),
    count_prenatal_exp = sum(prenatal_exp),
    prop_prenatal_exp = mean(prenatal_exp),
  ) %>%
  t %>%
  as_tibble(rownames = "Statistic")
colnames(buccal_table) <- c("Statistic", "Buccal Value")
```

```{r create-table-pbmc, warning=FALSE, message=FALSE}
pbmc_table <- pbmc_pheno_tbl %>%
  mutate(
    male = if_else(Sex == "M", 1, 0), # assume that a == man, b == women
    ever_smoker = if_else(Smoking == "a", 1, 0),
    prenatal_exp = if_else(Exposed == "yes", 1, 0)
  )  %>%
  summarise(
    participant_count = n(),
    mean_age = mean(Age),
    sd_age = sd(Age),
    count_male = sum(male),
    prop_male = mean(male),
    count_ever_smoker = sum(ever_smoker),
    prop_ever_smoker = mean(ever_smoker),
    count_prenatal_exp = sum(prenatal_exp),
    prop_prenatal_exp = mean(prenatal_exp),
  ) %>%
  t %>%
  as_tibble(rownames = "Statistic")
colnames(pbmc_table) <- c("Statistic", "PBMC Value")
```


```{r join_tables}
# print summary statistics
buccal_table %>%
  left_join(pbmc_table, by = "Statistic") %>%
  kable(digits = 2)
```

Unfortunately, there appears to be a few differences in the sample metadata. The
number of participants who have ever smoked and who were exposed to arsenic in
the  womb also differ by 1.

To identify where these metadata files diverge, we'll match the samples to their
study IDs. Also of note: there are two samples labeled "P24" and two samples
labeled "B24". This is likely a recording error; these participants have
different study IDs, and have different metadata. They are likely different
people.


```{r combine-data, warning=FALSE, message=FALSE}
# load the sample vs study ID key
key_tbl <- read_csv(
  "../../../arsenic-epigenetics-meta/Chile/study_vs_sample_id.csv"
)

# load the buccal sample metadata
blind_buccal_tbl <- read_csv(
  "../../../arsenic-epigenetics-meta/Chile/buccal/data/blind-metadata-buccal-AsChile.csv"
)

# rename the sample IDs and some columns
blind_buccal_tbl <- blind_buccal_tbl %>%
  filter(`Sample ID` != "QC") %>%
  mutate(
    sample_id = `Sample ID`,
    Exposed = `Exposed: Yes:a; No:b`
  ) %>%
  select(sample_id, Exposed, P3Q8_Age, P3Q9_Sex, P38Q96_Smoking) %>%
  left_join(key_tbl, by = "sample_id") %>%
  filter(!(sample_id == "B24" & study_id == "AC130101" & P3Q8_Age == 52)) %>%
  filter(!(sample_id == "B24" & study_id == "AC130016" & P3Q8_Age == 51))

# load the pbmc sample metadata
blind_pbmc_tbl <- read_csv(
  "../../../arsenic-epigenetics-meta/Chile/pbmcs/data/blind-metadata-pbmcs-AsChile.csv"
)

# rename the sample IDs and some columns
blind_pbmc_tbl <- blind_pbmc_tbl %>%
  filter(`Sample ID` != "QC") %>%
  mutate(
    sample_id = `Sample ID`,
  ) %>%
  select(sample_id, Exposed, P3Q8_Age, P3Q9_Sex, P38Q96_Smoking) %>%
  left_join(key_tbl, by = "sample_id") %>%
  filter(!(sample_id == "P24" & study_id == "AC130101" & P3Q8_Age == 52)) %>%
  filter(!(sample_id == "P24" & study_id == "AC130016" & P3Q8_Age == 51))

# join the tables based on sample_id
combo_metada_tbl <- blind_buccal_tbl %>%
  left_join(blind_pbmc_tbl, by = "study_id",
            suffix = c(".buccal", ".pbmc")) %>%
  arrange(study_id) %>%
  select(study_id, everything())

# print combined metadata table
combo_metada_tbl %>% kable()
```


This is bad news. The analyses performed with both the buccal cells and PBMCs
each contain 44 observations. However, there are only 40 unique study IDs. That
means that there are 4 participants in each study who has technical replicates.
These will likely need to be consolidated, and the analyses repeated. The list
of technical replicates are given below:

Buccal cells:

- Participant AC130004: B1 and B36
- Participant AC130037: B16 and B34
- Participant AC130044: B2 and B5
- Participant AC130079: B7 and B41

PBMCs:

- Participant AC130010: P17 and P29
- Participant AC130031: P3 and P46
- Participant AC130048: P5 and P10
- Participant AC130084: P4 and P15

This explains the difference in descriptive statistics found earlier: the
participants with technical replicates in the buccal cell data are not the
same as the participants with techincal replicates in the PBMCs data.

## Corrected Descriptive Statistics

In the table below, I've re-computed the descriptive statistics, filtering out
the duplicated participants from each study. They match now, and so the results
can be put into Table 1.

```{r true-descriptive-stats}
# buccal cells
cleaned_buccal_desc_stats <- blind_buccal_tbl %>%
  select(-sample_id) %>%
  distinct %>%
  mutate(
    male = if_else(P3Q9_Sex == "a", 1, 0),
    ever_smoker = if_else(P38Q96_Smoking == "a", 1, 0),
    prenatal_exp = if_else(Exposed == "a", 1, 0),
    P3Q8_Age = as.numeric(P3Q8_Age)
  ) %>%
  summarise(
    participant_count = n(),
    mean_age = mean(P3Q8_Age),
    sd_age = sd(P3Q8_Age),
    count_male = sum(male),
    prop_male = mean(male) * 100,
    count_ever_smoker = sum(ever_smoker),
    prop_ever_smoker = mean(ever_smoker) * 100,
    count_prenatal_exp = sum(prenatal_exp),
    prop_prenatal_exp = mean(prenatal_exp) * 100,
  ) %>%
  t %>%
  as_tibble(rownames = "Statistic")
colnames(cleaned_buccal_desc_stats) <- c("Statistic", "Buccal Value")

# PBMCs
cleaned_pbmc_desc_stats <- blind_pbmc_tbl %>%
  select(-sample_id) %>%
  distinct %>%
  mutate(
    male = if_else(P3Q9_Sex == "a", 1, 0),
    ever_smoker = if_else(P38Q96_Smoking == "a", 1, 0),
    prenatal_exp = if_else(Exposed == "a", 1, 0),
    P3Q8_Age = as.numeric(P3Q8_Age)
  ) %>%
  summarise(
    participant_count = n(),
    mean_age = mean(P3Q8_Age),
    sd_age = sd(P3Q8_Age),
    count_male = sum(male),
    prop_male = mean(male) * 100,
    count_ever_smoker = sum(ever_smoker),
    prop_ever_smoker = mean(ever_smoker) * 100,
    count_prenatal_exp = sum(prenatal_exp),
    prop_prenatal_exp = mean(prenatal_exp) * 100,
  ) %>%
  t %>%
  as_tibble(rownames = "Statistic")
colnames(cleaned_pbmc_desc_stats) <- c("Statistic", "PBMC Value")

# compare the two
cleaned_buccal_desc_stats %>%
  left_join(cleaned_pbmc_desc_stats, by = "Statistic") %>%
  kable(digits = 2)
```

## Post QC Results

After applying some quality control measures to the buccal cell samples, we
found that one of the samples was of low quality. It was therefore removed from
the analysis. The updated decsriptive statistics for the buccal cell samples
are computed below.

```{r updated-buccal}
post_qc_buccal_desc_stats <- blind_buccal_tbl %>%
  filter(sample_id != "B14") %>%
  select(-sample_id) %>%
  distinct %>%
  mutate(
    male = if_else(P3Q9_Sex == "a", 1, 0),
    ever_smoker = if_else(P38Q96_Smoking == "a", 1, 0),
    prenatal_exp = if_else(Exposed == "a", 1, 0),
    P3Q8_Age = as.numeric(P3Q8_Age)
  ) %>%
  summarise(
    participant_count = n(),
    mean_age = mean(P3Q8_Age),
    sd_age = sd(P3Q8_Age),
    count_male = sum(male),
    prop_male = mean(male) * 100,
    count_ever_smoker = sum(ever_smoker),
    prop_ever_smoker = mean(ever_smoker) * 100,
    count_prenatal_exp = sum(prenatal_exp),
    prop_prenatal_exp = mean(prenatal_exp) * 100,
  ) %>%
  t %>%
  as_tibble(rownames = "Statistic")
colnames(post_qc_buccal_desc_stats) <- c("Statistic", "Buccal Value")
post_qc_buccal_desc_stats %>% kable(digits = 2)
```

